Technique for \earlywarning\ detection of compact binary coalescence and its implications for multi-messenger astronomy

### 2011-06-28 introduction.tex
Converted some figures to eps
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1162 ccdd05d5-56c1-461a-bf98-85f6876523b4
tone down
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1167 ccdd05d5-56c1-461a-bf98-85f6876523b4

As a compact binary system loses energy to \GW{}s, its
orbital separation decays, leading to a runaway inspiral with the \GW{}
amplitude and frequency increasing until the system eventually merges.  It is
thought that if a \textsc{ns} is involved, it may become tidally disrupted near
the merger and fuel a bright electromagnetic (\EM{})
counterpart \citep{shibata:2007}. One way of obtaining such joint observations
is through the rapid identification and dissemination of candidate \GW{} events
for \EM{} followup.

In the first generation of ground-based laser interferometers, the \GW{}
community initiated a
project to send alerts when potential \GW{} transients were observed in order
to trigger followup observations by \EM{} telescopes.  The typical latencies
were 30--60 minutes \citep{HugheyGWPAW2011}, which was an important
achievement, but far too late to catch prompt \EM{} emission. We have the
ambition of reporting \GW{} candidates not minutes \emph{after} the merger, as
has already been accomplished, but seconds \emph{before}.

@@

Though compact binary coalescence (\CBC{}) events may be responsible for several
classes of \EM{} transients, \CBC{}s are in particular believed to be a
mechanism for short gamma-ray bursts (short \GRB{}s) \citep{Lee:2005, nakar07}.
In this scenario, prompt \EM{} emission arises as shells of
relativistically out-flowing matter collide in the inner shock.
The same inner shocks, or potentially reverse shocks, can produce a bright
accompanying optical flash \citep{Sari99}. Prompt emission is a probe into
the extreme initial conditions of the outflow, in contrast with afterglows,
which arise in the external shock with the local medium and are relatively
insensitive to initial conditions.

@@

In October 2010, \LIGO{}\footnote{\url{http://www.ligo.org/}} completed its sixth science run
(S6) and Virgo\footnote{\url{http://www.ego-gw.it/}} completed its third science run (VSR3).
While both \LIGO{} detectors and Virgo were operating, several all-sky detection
pipelines operated in a low-latency configuration to send astronomical alerts,
namely \textsc{mbta}, Coherent WaveBurst, and
Omega \citep{HugheyGWPAW2011, S6lowlatency}. The S6 analyses
achieved latencies of 30--60 minutes, which were dominated by a human vetting
process. Candidates were sent for \EM{} followup to several
telescopes; Swift, \textsc{rotse}, \textsc{tarot}, and Zadko \citep{kanner2008,
HugheyGWPAW2011} imaged some of the most likely sky locations.  \textsc{mbta} achieved
the best \GW{} trigger-generation latencies of 2--5 minutes.  We
assume that in the advanced detector era the vetting process will be automated,
so current \GW{} search methodology and telescope actuation would
dominate latency.

\begin{figure}[h]
\plotone{figures/snr_in_time.pdf}
\caption{\label{fig:earlywarning}Expected number of \textsc{ns}--\textsc{ns}
sources that will be detectable by Advanced \LIGO\ a given number of seconds before coalescence.
The heavy solid line is the most realistic yearly rate estimate.  The shaded
region represents the 5 to 95\% confidence interval arising from uncertainty
in predicted event rates \citep{Abadie:2010p10836}.}
\end{figure}
%
It has been observed that by examining the signal to noise ratio (\SNR{}) stream for
threshold crossings before the \GW{} signal leaves the detection band, it is
possible to trade some \SNR{} and sky localization accuracy for pre-merger
detection. Figure~\ref{fig:earlywarning} shows projected early detection rates
for \textsc{ns}--\textsc{ns} binaries in Advanced \LIGO\ assuming event rates
predicted in \citet{Abadie:2010p10836} and anticipated detector sensitivity for
the `zero detuning, high power' configuration described in \citet{ALIGONoise}.
The most realistic estimates indicate that at an \SNR\ threshold of 8, we will
observe a total of 40~events~yr$^{-1}$, but $\sim$10~yr$^{-1}$ will be
detectable within 10~s of merger and $\sim$5~yr$^{-1}$ will be detectable
within 25~s of merger. These rates are shown with their substantial uncertainties in Figure~\ref{fig:earlywarning}.

\begin{table}[h]
\caption{\label{table:sky-localization-accuracy}Horizon distance, \SNR\ at merger, and area of 90\% confidence at selected times before merger for sources with expected detection rates of 40, 10, 1, and 0.1~yr$^{-1}$.}
\begin{center}
\begin{tabular}{rrrrrrr}
\tableline\tableline
rate & horiz. & final & \multicolumn{4}{c}{$A$(90\%) (deg$^2$)} \\
\cline{4-7}
yr$^{-1}$ & (Mpc) & \SNR\ & 25 s & 10 s & 1 s & 0 s \\
\tableline
40\phd\phn & 445 & 8.0 & 15500 & 3100 & 200 & 9.6 \\
10\phd\phn & 280 & 12.7 & 6100 & 1200 & 78 & 3.8 \\
1\phd\phn & 130 & 27.4 & 1300 & 260 & 17 & 0.8 \\
0.1 & 60 & 58.9 & 280 & 56 & 3.6 & 0.2 \\
\tableline
\end{tabular}
\end{center}
\end{table}
%
\begin{figure}[h]
\plotone{figures/loc_in_time.pdf}
\caption{\label{fig:sky-localization-accuracy}Area of the 90\% confidence
region as a function of time for a sources with anticipated detection rates
of 40, 10, 1, and 0.1~yr$^{-1}$. The heavy dot indicates the
time at which the accumulated \SNR\ exceeds a threshold of 8.}
\end{figure}

\EM{} followup requires estimating the location of the \GW{} source. The
localization uncertainty is dominated by uncertainty in the time of arrival of
the \GW{}s, which is determined by the signal's effective bandwidth and \SNR{}
\citep{Fairhurst2009}. Table~\ref{fig:earlywarning} and Figure~\ref{fig:sky-localization-accuracy} show the estimated
minimum 90\% confidence area versus time of the loudest
coalescence events detectable by Advanced \LIGO{} and Advanced Virgo.  Once per year,
we expect to observe an event with a final \SNR\ of 27.4 whose location can be
constrained to about 1300~deg$^2$ (3.1\% of the sky) 25~s before merger, 260~deg$^2$
(0.63\% of the sky) 10~s before merger, and 0.82~deg$^2$ (0.0020\% of the sky) at merger.

It is infeasible to search hundreds of square degrees for a prompt
counterpart. For
reference, \textsc{lsst} should see first light around the same time as
Advanced \LIGO{} and will have an unparalleled field of view of 9.6 deg$^2$
\citep{2008arXiv0805.2366I}.  However, if we can take galactic luminosities
to accurately reflect mass and thus be a proxy for the \CBC{} event rate, it is
possible to reduce the localization uncertainty by only looking at galaxies from a
catalog that lie near the sky location and luminosity distance estimate from the \GW{}
signal~\citep{galaxy-catalog} as was done in S6/VSR3.
Within the expected Advanced \LIGO{} \textsc{ns}--\textsc{ns} horizon
distance, the number of galaxies that can produce a given signal
amplitude is much larger than the number for initial \LIGO{} and thus the
catalog will not be as useful for downselecting pointings generally. However,
our exceptional sources will necessarily be extremely nearby. Within this
reduced volume there will be fewer galaxies to consider for a given
candidate and catalog completeness will be less of a concern.
This should allow us to shrink the 90\% confidence area substantially.

\subsection{Latency in \CBC{} search algorithms}

Realizing advance detection of \CBC{}s will
require striking a balance between latency and throughput. \CBC{} searches
consist of banks of matched filters, or
cross-correlations between the data stream and a bank of nominal ``template''
signals.  There are many different implementations of matched filters, but most
have high throughput at the cost of high latency, or low latency at the cost of
low throughput.  The former are epitomized by the overlap-save algorithm~%
\citep{numerical-recipes-chapter-13} for frequency domain (\FD) convolution,
currently the preferred method in \GW{}
searches.  The most obvious example of the latter is the time domain
(\TD) convolution, which is latency-free.  However, its
computational complexity is quadratic in the length of the templates, so it is
prohibitively expensive for long templates.

Fortunately, the morphology of inspiral signals can be exploited to offset some
of the computational complexity of low-latency algorithms.  First, the signals
evolve slowly in frequency, so that they can be broken into contiguous
band-limited time intervals and processed at possibly lower sample rates.
Second, inspiral filter banks consist of highly similar templates, admitting
methods such as singular value decomposition (\SVD{}) to reduce the number of
templates \citep{Cannon:2010p10398}. We will use both aspects to demonstrate that a very
low latency analysis with advance detection of compact binary sources is
possible with current computing resources.  Assuming other technical sources of
latency can be reduced significantly, this should allow the possibility for
prompt alerts to be sent to the astronomical community.

>>>>

As a compact binary system loses energy to gravitational waves (\GW{}s), its
orbital separation decays, leading to a runaway inspiral with the \GW{}
amplitude and frequency increasing until the system eventually merges.  It is
thought that if a \textsc{ns} is involved, it may become tidally disrupted near
the merger and fuel a bright electromagnetic (\EM{}) counterpart
\citep{shibata:2007}.  The gravitational wave signal is in principle detectable
before merger.  Effort from both the \GW{} and astronomy communities may make
it possible to use \GW{} observations as an early warning trigger for \EM{}
followup.  We have the ambition of reporting GW candidates not minutes after
the merger, as has already been accomplished, but seconds before.  We discuss a
real-time filtering algorithm for compact binary coalescence waveforms, which
is one aspect of the work required to make this possible.  We also present the
prospects for advanced \GW\ detectors and discuss other areas of work that
would be required for rapid analysis.

In the first generation of ground-based laser interferometers, the \GW{}
community initiated a project to send alerts when potential \GW{} transients
were observed in order to trigger followup observations by \EM{} telescopes.
The typical latencies were 30 minutes \citep{HugheyGWPAW2011}, which was an
important achievement, but probably too late to catch the potential prompt
\EM{} emission.

@@

Though compact binary coalescence (\CBC{}) events may be responsible for
several classes of \EM{} transients, \CBC{}s are in particular believed to be a
mechanism for short gamma-ray bursts (short \GRB{}s) \citep{Lee:2005, nakar07}.
In this scenario, prompt gamma-ray emission arises on the accretion timescale
of a central compact object formed after the merger, 0.1--1.0\,s
\citep{Janka1999}.  Interestingly, roughly 1/4 -- 1/2 of observed short \GRB{}s
also exhibit extended emission $\sim100$\,s.  Potential explanations for the
emission are delayed fall-back accretion \citep{Metzger2010} caused by
$r$-process heating or the formation of a proto-magnetar that interacts with
ejecta \citep{Bucciantini2011}.  In the case of the proto-magnetar, it might be
formed from accretion induced collapse, or the merger of two neutron stars.  An
accompanying \GW{} observation would confirm the latter.

@@

In October 2010, \LIGO{}\footnote{\url{http://www.ligo.org/}} completed its
sixth science run (S6) and Virgo\footnote{\url{http://www.ego-gw.it/}}
completed its third science run (VSR3).  While both \LIGO{} detectors and Virgo
were operating, several all-sky detection pipelines operated in a low-latency
configuration to send astronomical alerts, namely \textsc{mbta}, Coherent
WaveBurst, and Omega \citep{HugheyGWPAW2011, S6lowlatency}. The S6 analyses
achieved latencies of 30--60 minutes, which were dominated by a human vetting
process. Candidates were sent for \EM{} followup to several telescopes; Swift,
\textsc{rotse}, \textsc{tarot}, and Zadko \citep{kanner2008, HugheyGWPAW2011}
imaged some of the most likely sky locations.  \textsc{mbta} achieved the best
\GW{} trigger-generation latencies of 2--5 minutes.  
%We assume that in the
%advanced detector era the vetting process will be automated, so current \GW{}
%search methodology and telescope actuation would dominate latency.

\begin{figure}[h]
\plotone{figures/snr_in_time}
\caption{\label{fig:earlywarning}Expected number of \textsc{ns}--\textsc{ns}
sources that could be detectable by Advanced \LIGO\ a given number of seconds
before coalescence.  The time before coalescence that a source could be
detectable is a lower bound that does not include any latency inducing aspects
of a real analysis.  The heavy solid line is the most realistic yearly rate
estimate.  The shaded region represents the 5 to 95\% confidence interval
arising from uncertainty in predicted event rates \citep{Abadie:2010p10836}.}
\end{figure}
%
Examining the signal to noise ratio (\SNR{}) stream for threshold crossings
before the \GW{} signal leaves the detection band makes it possible to trade
some \SNR{} and sky localization accuracy for pre-merger detection.
Figure~\ref{fig:earlywarning} shows projected early detectability rates for
\textsc{ns}--\textsc{ns} binaries in Advanced \LIGO\ assuming event rates
predicted in \citet{Abadie:2010p10836} and anticipated detector sensitivity for
the `zero detuning, high power' configuration described in \citet{ALIGONoise}.
The most realistic estimates indicate that at an \SNR\ threshold of 8, we will
observe a total of 40~events~yr$^{-1}$. $\sim$10~yr$^{-1}$ will be detectable
within 10~s of merger and $\sim$5~yr$^{-1}$ will be detectable within 25~s of
merger if analysis can proceed with near zero latency. These rates are shown
with their substantial uncertainties in Figure~\ref{fig:earlywarning}.

We emphasize that Figure~\ref{fig:earlywarning} provides only the lower bound
for early detection and should serve as a baseline to understand the potential
of early warning \GW{} searches in a ideal world.  Real analysis will have
sources of latency that are unavoidable meaning that the actual detection will
occur with some delay.

\begin{table}[h]
\caption{\label{table:sky-localization-accuracy}Horizon distance, \SNR\ at
merger, and area of 90\% confidence at selected times before merger for sources
with expected detectability rates of 40, 10, 1, and 0.1~yr$^{-1}$.}
\begin{center}
\begin{tabular}{rrrrrrr}
\tableline\tableline
rate & horiz. & final & \multicolumn{4}{c}{$A$(90\%) (deg$^2$)} \\
\cline{4-7}
yr$^{-1}$ & (Mpc) & \SNR\ & 25 s & 10 s & 1 s & 0 s \\
\tableline
40\phd\phn & 445 & 8.0 & 15500 & 3100 & 200 & 9.6 \\
10\phd\phn & 280 & 12.7 & 6100 & 1200 & 78 & 3.8 \\
1\phd\phn & 130 & 27.4 & 1300 & 260 & 17 & 0.8 \\
0.1 & 60 & 58.9 & 280 & 56 & 3.6 & 0.2 \\
\tableline
\end{tabular}
\end{center}
\end{table}
%
\begin{figure}[h]
\plotone{figures/loc_in_time}
\caption{\label{fig:sky-localization-accuracy}Area of the 90\% confidence
region as a function of time before coalescence for sources with anticipated
detectability rates of 40, 10, 1, and 0.1~yr$^{-1}$. The heavy dot indicates
the time at which the accumulated \SNR\ exceeds a threshold of 8.}
\end{figure}

Before discussing realistic latencies, it is important to note that \EM{}
followup requires estimating the location of the \GW{} source. The localization
uncertainty is dominated by uncertainty in the time of arrival of the \GW{}s,
which is determined by the signal's effective bandwidth and \SNR{}
\citep{Fairhurst2009}.  Table~\ref{fig:earlywarning} and
Figure~\ref{fig:sky-localization-accuracy} show the estimated minimum 90\%
confidence area versus time of the loudest coalescence events detectable by
Advanced \LIGO{} and Advanced Virgo.  Once per year, we expect to observe an
event with a final \SNR\ of $\sim 27$ whose location can be constrained to about
1300~deg$^2$ (3.1\% of the sky) using only the waveform 25~s before merger,
260~deg$^2$ (0.63\% of the sky) using the waveform 10~s before merger, and
0.82~deg$^2$ (0.0020\% of the sky) using the full \GW\ waveform.

It is infeasible to search hundreds of square degrees for a prompt counterpart.
For reference, \textsc{lsst} should see first light around the same time as
Advanced \LIGO{} and will have an unparalleled field of view of 9.6 deg$^2$
\citep{2008arXiv0805.2366I}.  However, 
%if we can take galactic luminosities to
%accurately reflect mass and thus be a proxy for the \CBC{} event rate, 
it is possible to reduce the localization uncertainty by only looking at
galaxies from a catalog that lie near the sky location and luminosity distance
estimate from the \GW{} signal~\citep{galaxy-catalog} as was done in S6/VSR3.
Within the expected Advanced \LIGO{} \textsc{ns}--\textsc{ns} horizon distance,
the number of galaxies that can produce a given signal amplitude is much larger
than the number for initial \LIGO{} and thus the catalog will not be as useful
for downselecting pointings generally. However, exceptional \GW{} sources will
necessarily be extremely nearby. Within this reduced volume there will be fewer
galaxies to consider for a given candidate and catalog completeness will be
less of a concern.  This should reduce the 90\% confidence area substantially.
Also future \GW{} detectors, should they be built, will improve the \SNR\ and
likewise the error box.

\subsection{Latency in \CBC{} searches}

Here we provide a brief list of technical latencies associated with \CBC\
analysis in S6, VSR3 in chronological order \citep{HugheyGWPAW2011}.
%
\begin{itemize}
%
\item {\it Data calibration, quality control and distribution ($\sim
2$\,minutes):}  This is the time it took to calibrate the detector data, assess
its quality and distribute it to a central location for analysis.  Improvements
in other areas could make this the dominating source of latency. If possible
work should be done to improve this.
%
\item {\it Trigger production ($2-5$\,minutes):} This was an impressive latency
and was certainly sub-dominant in the S6, VSR3 analysis.  However, even if no
other sources of latency exist it is too long to catch prompt or extended
emission.  It is also more challenging for searches with advanced detectors
because the templates needed to filter the data are about 10 times longer.  We
will focus on reducing the latency of the trigger production step in this work.
%
\item {\it Event gathering, downselection and pointing calculations ($2-3$\,
minutes):} This was the time required to gather the event parameters and sky
localization from the various online analyses, downselect the events and
calculate pointings.  If other sources of latency were improved this could
dominate.  If possible work should be done to improve this.
%
\item {\it Human vetting of the trigger quality ($10-20$\,minutes):}  This was
by far the largest latency during the S6, VSR3 low latency analysis and will
hopefully be automated by the advanced detector era.  We therefore give it no
more consideration here.
%
\end{itemize}
%

In addition to the latencies described above, which may be overcome, there are
fundamental latencies associated with data acquisition and transfer that cannot
be overcome.  For example, the \LIGO{} data acquisition records science data in
1/16~s blocks~\citep{Bork2001}. Data must also be aggregated from all of the
\GW\ observatories using global computer networks capable of high bandwidth but
perhaps only modest latency.  This could introduce an additional latency of
$\sim$100~ms for a total of $\sim 1/4$\,s of unavoidable latency.

This work will focus on reducing the latency of trigger production.  Realizing
advance detection of \CBC{}s will require striking a balance between latency
and throughput of filtering \GW\ wave data. \CBC{} searches consist of banks of
matched filters, or cross-correlations between the data stream and a bank of
nominal ``template'' signals.  There are many different implementations of
matched filters, but most have high throughput at the cost of high latency, or
low latency at the cost of low throughput.  The former are epitomized by the
overlap-save algorithm~% \citep{numerical-recipes-chapter-13}
for frequency domain (\FD) convolution, currently the preferred method in \GW{}
searches.  The most obvious example of the latter is the time domain (\TD)
convolution, which is latency-free.  However, its computational complexity is
quadratic in the length of the templates, so it is prohibitively expensive for
long templates.

Fortunately, the morphology of inspiral signals can be exploited to offset some
of the computational complexity of low-latency algorithms.  First, the signals
evolve slowly in frequency, so that they can be broken into contiguous
band-limited time intervals and processed at possibly lower sample rates.
Second, inspiral filter banks consist of highly similar templates, admitting
methods such as singular value decomposition (\SVD{}) to reduce the number of
templates \citep{Cannon:2010p10398}. We will use both aspects to demonstrate
that a very low latency detection statistic is possible with current computing
resources.  Assuming the other technical sources of latency can be reduced
significantly, this should allow the possibility for prompt alerts to be sent
to the astronomical community.
